---
title: "SVM Model"
author: "Hanan Ali"
date: "2024-12-05"
output: html_document
---

### Loading libraries
```{r}
# Load libraries
library(e1071)
library(caret)
library(dplyr)
library(ggplot2)
library(tidyr)
library(readxl)
```

### Pre-processing Data 
```{r}
# Import the data
diabetes_data <- read_excel("diabetes_binary_health_indicators_BRFSS2015.xlsm")

# Convert variables to factors
diabetes_data$Diabetes_binary <- factor(diabetes_data$Diabetes_binary, levels = c(0, 1), labels = c("No Diabetes", "Prediabetes or Diabetes"))
diabetes_data$HighBP <- factor(diabetes_data$HighBP, levels = c(0, 1), labels = c("No", "Yes"))
diabetes_data$HighChol <- factor(diabetes_data$HighChol, levels = c(0, 1), labels = c("No", "Yes"))
diabetes_data$CholCheck <- factor(diabetes_data$CholCheck, levels = c(0, 1), labels = c("No", "Yes"))
diabetes_data$Smoker <- factor(diabetes_data$Smoker, levels = c(0, 1), labels = c("No", "Yes"))
diabetes_data$Stroke <- factor(diabetes_data$Stroke, levels = c(0, 1), labels = c("No", "Yes"))
diabetes_data$HeartDiseaseorAttack <- factor(diabetes_data$HeartDiseaseorAttack, levels = c(0, 1), labels = c("No", "Yes"))
diabetes_data$PhysActivity <- factor(diabetes_data$PhysActivity, levels = c(0, 1), labels = c("No", "Yes"))
diabetes_data$Fruits <- factor(diabetes_data$Fruits, levels = c(0, 1), labels = c("No", "Yes"))
diabetes_data$Veggies <- factor(diabetes_data$Veggies, levels = c(0, 1), labels = c("No", "Yes"))
diabetes_data$HvyAlcoholConsump <- factor(diabetes_data$HvyAlcoholConsump, levels = c(0, 1), labels = c("No", "Yes"))
diabetes_data$AnyHealthcare <- factor(diabetes_data$AnyHealthcare, levels = c(0, 1), labels = c("No", "Yes"))
diabetes_data$NoDocbcCost <- factor(diabetes_data$NoDocbcCost, levels = c(0, 1), labels = c("No", "Yes"))
diabetes_data$DiffWalk <- factor(diabetes_data$DiffWalk, levels = c(0, 1), labels = c("No", "Yes"))
diabetes_data$Sex <- factor(diabetes_data$Sex, levels = c(0, 1), labels = c("Male", "Female"))
diabetes_data$GenHlth <- factor(diabetes_data$GenHlth, levels = 1:5, labels = c("Excellent", "Very Good", "Good", "Fair", "Poor"))

# Convert Education, and Income to factors with meaningful level names
diabetes_data$Education <- factor(diabetes_data$Education, levels = 1:6, labels = c("Never attended school", "Elementary", "Some high school", "High school graduate", "Some college", "College graduate"))
diabetes_data$Income <- factor(diabetes_data$Income, levels = 1:8, labels = c("Less than $10,000", "$10,000-$14,999", "$15,000-$19,999", "$20,000-$24,999", "$25,000-$34,999", "$35,000-$49,999", "$50,000-$74,999", "$75,000 or more"))

# Remove BMI observations greater than 50 and store in a new dataset
diabetes_data <- diabetes_data %>%
  filter(BMI < 50)

# Select relevant variables
diabetes_data_svm <- diabetes_data %>%
  select(Diabetes_binary, HighBP, HighChol, CholCheck, HeartDiseaseorAttack, HvyAlcoholConsump, GenHlth, Sex, Age)

# Convert target variable to a factor
diabetes_data_svm$Diabetes_binary <- as.factor(diabetes_data_svm$Diabetes_binary)

# Scale numeric variables
numeric_vars <- c("Age")  # Add more numeric variables if needed
diabetes_data_svm <- diabetes_data_svm %>%
  mutate(across(all_of(numeric_vars), scale))

```


### Split Data into Training and Testing Sets
```{r}
# Set seed for reproducibility
set.seed(123)

# Create training (70%) and testing (30%) datasets
train_index <- createDataPartition(diabetes_data_svm$Diabetes_binary, p = 0.7, list = FALSE)
train_data <- diabetes_data_svm[train_index, ]
test_data <- diabetes_data_svm[-train_index, ]
```


### Create an SVM Model with a radial kernel with cost=1 (default value). The radial kernel is capable of modeling nonlinear boundaries in the data and is useful when the relationship between the predictors and the target variable isn't linear.
```{r}
# Train SVM model with radial kernel and cost = 1
svm_model <- svm(Diabetes_binary ~ ., 
                 data = train_data, 
                 kernel = "radial", 
                 type = "C-classification")
```


### Tuning hyperparameters:
A low cost encourages a simpler model that may misclassify some training points but could generalize better. A high cost penalizes misclassification more, fitting the training data more closely but increasing the risk of overfitting. On the other hand, a low gamma indicates that the SVM looks at data points far away to decide the classification of a given point. A high gamma indicates that the SVM focuses only on points very close to the one it’s trying to classify.

I'm going to use grid search with cross-validation to find the best value for cost and gamma. 
```{r}
# Train the SVM model with specified cost and gamma
svm_model_fixed <- svm(
  Diabetes_binary ~ ., 
  data = train_data, 
  kernel = "radial", 
  type = "C-classification", #using classification instead of regression because better choice for predicting binary outcomes
  cost = 1, 
  gamma = 0.1
)

# Print the model summary
print(summary(svm_model_fixed))
```


### Looking at model's performance
```{r}
# Predict on the test set
predictions <- predict(svm_model_fixed, newdata = test_data)

# Confusion matrix
conf_matrix <- confusionMatrix(predictions, test_data$Diabetes_binary)
print(conf_matrix)

# Accuracy
accuracy <- conf_matrix$overall["Accuracy"]
cat("Model Accuracy:", accuracy, "\n")
```
Sensitivity is high, meaning the model does a good job at identifying those diabetes. Specificity is very low, suggesting that the model fails to identify people without diabetes correctly.

The model has a bias towards predicting diabetes, as reflected in the high senstivity and low specificity. The model seems to be over-predicting diabetes and under-predicting no diabetes, resulting in a high accuracy but poor balanced performance.






### Setting up another SVM model that trains an SVM model with a linear kernel, using repeated cross-validation and testing different values of the cost parameter as defined in the tuning grid. The best cost value is selected based on model performance.
```{r}
# Set up repeated cross-validation
# It is configuring the cross-validation process to split the dataset into folds, train the model on some folds, and test it on others repeatedly. The training and test sets are created within each fold during the cross-validation process.
set.seed(100)
tr_control <- trainControl(method = "repeatedcv", 
                           number = 5,  # Number of folds
                           repeats = 10)  # Number of repetitions

# Define a tuning grid for the cost parameter
tune_grid <- expand.grid(cost = exp(seq(-5, 3, length.out = 1)))#this generates a sequence of 1 value between -5 and 3
#ANY VALUE MORE THAN 1 IN LENGTH.OUT WILL LEAD TO MODEL CRASHING 

# Train the SVM model using the caret package
svm_caret <- train(
  Diabetes_binary ~ .,  # Formula with target and predictors
  data = train_data,    # Training data
  method = "svmLinear2",  # SVM with linear kernel and custom cost tuning
  tuneGrid = tune_grid,  # Grid for tuning the cost parameter
  trControl = tr_control # Repeated CV settings
)

# Output the best cost parameter
cat("Best Cost Parameter (C):", svm_caret$bestTune$cost, "\n")

# Print the model summary
print(svm_caret)
```


```{r}
# Make predictions on the test data using the trained SVM model
pred.class <- predict(object = svm_caret, 
                      newdata = test_data, 
                      type = "raw")  # "raw" to get class predictions, not probabilities

# Calculate the error matrix using klaR package
err <- klaR::errormatrix(true = test_data$Diabetes_binary, 
                         predicted = pred.class, 
                         relative = TRUE)

# Round the error matrix to 3 decimal places and print it
round(err, 3)

# To extract the test accuracy from the error matrix:
accuracy <- sum(diag(err))  # Sum of the diagonal elements (correct predictions)
cat("Test Accuracy:", accuracy, "\n")

```


